---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(scrooge)
```

# Statistical modelling of heterogeneous citation networks

An R package containing utilities useful for the author's doctoral research in bibliometrics.

Install and load the package using

```r
devtools::install_github('Selbosh/scrooge')
library(scrooge)
```

## Verify if two numbers are approximately equal

```{r}
1 %=% 1.0000000000000000001
pi %=% 3.14159
```

## Bibliometric datasets

```{r}
head(articles)
```

## Journal ranking metrics

```{r}
head(names(ILSR(citations, sort = TRUE)))
head(names(PageRank(citations, sort = TRUE)))
head(names(Scroogefactor(citations, sort = TRUE)))
```

For comparison, here are the results of analysis of the same data using the `BradleyTerry2` package and the Bradley--Terry model:

```{r}
head(names(BTscores(citations)))
cor(BTscores(citations), ILSR(citations))
```

